---
title: 'Project 2'
author: "Akshay_Gupta_36-600_Project_2"
output:
  html_document:
  toc: false
  theme: spacelab
---

### Load necessary libraries
```{r}
suppressPackageStartupMessages(library(tidyverse))  # For data manipulation
suppressPackageStartupMessages(library(corrplot))   # For correlation plots
suppressPackageStartupMessages(library(car))        # For variance inflation factor (VIF) analysis
suppressPackageStartupMessages(library(leaps))      # For best subset selection
suppressPackageStartupMessages(library(glmnet))     # For Ridge and LASSO models
suppressPackageStartupMessages(library(caret))      # For data partitioning and cross-validation
```

### Load and Inspect the Data stellar_temperature.csv
```{r}
star_data <- read.csv("stellar_temperature.csv")

str(star_data)
summary(star_data)
colSums(is.na(star_data)) # no 'na' values
```

### Exploratory Data Analysis (EDA)
```{r}
# Visualize outliers in the target variable using a boxplot
ggplot(star_data, aes(y = teff)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of teff") +
  theme_minimal()

# Calculate Q1 (25th percentile) and Q3 (75th percentile)
Q1 <- quantile(star_data$teff, 0.25)
Q3 <- quantile(star_data$teff, 0.75)

# Calculate the IQR
IQR_value <- IQR(star_data$teff)

# Define bounds for identifying outliers
lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

# Create a new dataset without outliers
star_data_no_outliers <- star_data[star_data$teff >= lower_bound & star_data$teff <= upper_bound, ]

# Visualize the boxplot of teff without outliers
ggplot(star_data_no_outliers, aes(y = teff)) +
  geom_boxplot(fill = "maroon") +
  labs(title = "Boxplot of teff (Without Outliers)") +
  theme_minimal()

# Histogram for teff
hist(star_data$teff, main = "Histogram of teff", xlab = "teff", breaks = 20, col = "hotpink", border = "black")

# Apply a log transformation to teff for normalization
log_teff <- log(star_data$teff)

# Histogram for log(teff)
hist(log_teff, main = "Histogram of log(teff)", xlab = "log(teff)", breaks = 20, col = "deepskyblue", border = "black")

# Generate a correlation plot to visualize relationships between variables
cor_matrix <- cor(star_data) # Calculate correlation matrix
corrplot(cor_matrix, method = "circle") # Display correlation plot
```
The distribution of the target variable, teff is not normal and is right skewed.
I applied a log transform on teff to normalize as shown in the plots. But we will use the default data to train our models as the log transformed data is still not normally distributed. I removed some outliers to try and make the data more normally distributed.
For the correlation plot, strong positive correlations are indicated by dark blue circles such as those involving variables g_mag, b_mag, and r_mag. This suggests these variables are collinear.
Negative correlations are indicated by dark red circles, such as those involving br_col and teff, indicate strong negative correlations, which could also contribute to multicollinearity if these variables are used together in a model.

### Splitting the Data
```{r}
# Set a random seed for reproducibility of results
set.seed(123) 

# Create a random sample of row indices for the training set
# The sample size is 70% of the total number of rows in  star_data
train_index <- sample(seq_len(nrow(star_data)), size = 0.7 * nrow(star_data))

# Create the training data set using the sampled indices
train_data <- star_data[train_index, ]

# Create the test data set by excluding the training indices from the star_data (30%)
test_data <- star_data[-train_index, ]

# Print the size of the training set
cat("Training set size:", nrow(train_data), "\n")

# Print the size of the test set
cat("Test set size:", nrow(test_data), "\n")
```
I did a 70%-30% split for the training and test data since the dataset size is large (10000 rows)

### Analysis with Linear Regression
```{r}
# Fit linear regression model using teff as the response variable
lm_model <- lm(teff ~ . - teff, data = train_data)  # Exclude teff from predictors

# Display model summary
summary(lm_model)

# Generate predictions for both training and test sets
train_preds <- predict(lm_model, newdata = train_data)  # Predictions on training data
test_preds <- predict(lm_model, newdata = test_data)    # Predictions on test data

# Calculate MSE and RMSE for the test set using original teff
mse <- mean((test_data$teff - test_preds)^2)  # Mean Squared Error
rmse <- sqrt(mse)  # Root Mean Squared Error

# Print MSE and RMSE values
print(paste("MSE for the test set:", mse))
print(paste("RMSE for the test set:", rmse))

# Create a data frame for ggplot
prediction_data <- data.frame(
  Observed = test_data$teff,
  Predicted = test_preds
)

# Plot Predicted vs Observed values using ggplot
ggplot(prediction_data, aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.6) +  # Add points with some transparency
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "solid") +  # Diagonal line for perfect prediction
  labs(title = "Predicted vs Observed teff",
       x = "Observed teff",
       y = "Predicted teff") +
  theme_minimal()
```
The linear fit seems to be good, especially between 4000 and 7000 teff. However, the MSE and RSME values of 196969.1 and 443.8 are very high in relation to the teff values. However, we will try Ridge and LASSO modeling to try and get better better models.
An adjusted R^2 value of 0.6549 is not great, indicating that the linear model explains approximately 65.49% of the variance in the response variable.

### Distribution of Model Residuals
```{r}
# Generate predictions on the training data
preds <- predict(lm_model, newdata = train_data)  # Predictions from the linear model

# Calculate residuals in the original scale
original_residuals <- train_data$teff - preds

# Ensure residuals are in a data frame
residuals_df <- data.frame(original_residuals)

# Plotting the histogram of residuals in original scale
ggplot(data = residuals_df, aes(x = original_residuals)) +
  geom_histogram(binwidth = 100, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Residuals (Original Scale)",
       x = "Residuals (teff)",
       y = "Frequency") +
  theme_minimal()

# Calculate the residuals for normality test
# Perform the K-S test on the original residuals
ks_test_original <- ks.test(original_residuals, "pnorm", mean = mean(original_residuals), sd = sd(original_residuals))
print(ks_test_original)

# Interpret the p-value
if (ks_test_original$p.value < 0.05) {
  print("Reject the null hypothesis: Residuals are not normally distributed.")
} else {
  print("Fail to reject the null hypothesis: Residuals appear to be normally distributed.")
}
```
When I performed a Kolmogorov-Smirnov normality test on residuals with no transformation on the target variable, teff, I rejected the null hypothesis and the residuals were not normally distributed.

### Homoscedasticity vs. Heteroscedasticity (residuals versus the predicted test-set response values)
```{r}
# Generate predictions for the test set
predicted_values <- predict(lm_model, test_data)

# Calculate residuals
residuals <- test_data$teff - predicted_values

# Create a data frame for plotting
residuals_df <- data.frame(predicted = predicted_values, residuals = residuals)

# Plot residuals versus predicted values
ggplot(data = residuals_df, aes(x = predicted, y = residuals)) +
  geom_point(color = "blue", alpha = 0.5) +
  labs(title = "Residuals vs. Predicted Values",
       x = "Predicted Values (teff)",
       y = "Residuals") +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red")  # Add horizontal line at y = 0
```
The plot of the residuals versus the predicted test-set response values suggests that the variance around the regression line is σ2 and is constant (homoscedasticity). The scatter points hover around y=0.
The absence of a clear pattern or systematic structure in the residuals suggests that the model fits the data well. If there were patterns, such as curves or trends, it might indicate model inadequacy or a need for transformation.

### Variance Inflation Factors (VIF) to Detect Multicollinearity
```{r}
# Check for aliased coefficients in the initial model
alias_info <- alias(lm_model)  # Assuming lm_model is already defined with teff as response
print(alias_info)

# Inspect the correlation matrix, excluding 'teff'
cor_matrix <- cor(train_data[, -which(names(train_data) == "teff")])  # Only exclude teff
print(cor_matrix)

# Identify highly correlated pairs (greater than 0.75 but not equal to 1)
high_cor <- which(abs(cor_matrix) > 0.75 & abs(cor_matrix) < 1, arr.ind = TRUE)
print(high_cor)

# Remove a highly correlated variable (e.g., 'b_mag')
train_data_reduced <- train_data[, !names(train_data) %in% c("b_mag")]

# Fit the linear model again with the reduced dataset
lm_model_reduced <- lm(teff ~ ., data = train_data_reduced)

# Check for aliased coefficients again in the reduced model
alias_info_reduced <- alias(lm_model_reduced)
print(alias_info_reduced)

# Check VIF for the reduced model
vif_values <- vif(lm_model_reduced)
print("VIF values:")
print(vif_values)
```
b_mag, r_mag, and g_mag have strong correlations with each other and have multicollinearity.

### Best-Subset Selection Analysis
```{r}
# Perform best subset selection for potential model improvements
best_subset <- regsubsets(teff ~ ., data = train_data, nvmax = 5)
subset_summary <- summary(best_subset)
print(subset_summary)
```
### Fitting and MSE Calculation
```{r}
# Fit the best model
best_model_formula <- teff ~ B + br_col + r_mag + pmra + parallax
best_model <- lm(best_model_formula, data = train_data)

# Make predictions on the test data
bss_predictions <- predict(best_model, newdata = test_data)

# Calculate MSE
bss_mse <- mean((test_data$teff - bss_predictions)^2)
print(paste("MSE for the best subset model:", bss_mse))

# Calculate R^2 for the best model
bss_r2 <- summary(best_model)$r.squared
print(paste("R^2 for the best subset model:", bss_r2))

# Fit the full model for comparison
full_model <- lm(teff ~ ., data = train_data)  # This includes all predictors
full_predictions <- predict(full_model, newdata = test_data)

# Calculate MSE for the full model
full_mse <- mean((test_data$teff - full_predictions)^2)
print(paste("MSE for the full model:", full_mse))

# Calculate R^2 for the full model
full_r2 <- summary(full_model)$r.squared
print(paste("R^2 for the full model:", full_r2))
```
The full model has a lower MSE value compared to the best subset model. The full model also has a higher R^2 value compared to the best subset model.

### PCA Analysis and Biplot (for Multicollinearity Mitigation)
```{r}
# Conduct PCA on predictors, excluding 'teff'
pca_data_for_analysis <- train_data %>% select(-teff) %>% 
  select_if(is.numeric)  # Ensure we only use numeric columns

# Check the structure of the data being passed to PCA
print("Structure of PCA data:")
print(str(pca_data_for_analysis))

# Scale the training data
training_means <- colMeans(pca_data_for_analysis)
training_sds <- apply(pca_data_for_analysis, 2, sd)

pca_data_scaled <- scale(pca_data_for_analysis, center = training_means, scale = training_sds)

# Conduct PCA
pca_result <- prcomp(pca_data_scaled, scale. = FALSE)  # No need to scale again

# Check PCA results
summary(pca_result)

# Extract important variables from PC1 and PC2 loadings
loadings <- abs(pca_result$rotation[, 1:2])
important_vars <- rownames(loadings)[apply(loadings, 1, max) > 0.5]  # Threshold for selection

# Create a data frame for plotting
pca_data <- data.frame(PC1 = pca_result$x[, 1], PC2 = pca_result$x[, 2])

# Create the PCA biplot
pca_plot <- ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 2, alpha = 0.5, color = 'deepskyblue') +  # Increase point size and transparency
  geom_text(data = pca_result$rotation[important_vars, ], 
            aes(x = PC1, y = PC2, label = rownames(pca_result$rotation[important_vars, ])), 
            color = "black", size = 4, nudge_y = 0.02) +  # Nudge labels slightly to avoid overlap
  geom_segment(data = pca_result$rotation[important_vars, ], 
               aes(x = 0, y = 0, xend = PC1 * 2, yend = PC2 * 2),  # Scale arrows for visibility
               arrow = arrow(length = unit(0.2, "inches")), 
               color = "red") +  # Add arrows for loadings
  xlim(-10, 10) + ylim(-7.5, 7.5) +
  ggtitle("PCA Biplot (Top Influential Variables)") +
  theme_minimal() +  # Use a minimal theme for better aesthetics
  theme(plot.title = element_text(hjust = 0.5))  # Center the title

# Print the plot to ensure it appears in the output
print(pca_plot)

# Save the plot with specified dimensions
ggsave("pca_biplot.png", plot = pca_plot, width = 10, height = 10)
```

### PCA Regression Analysis
```{r}
# No. of PCs to keep
explained_variance <- summary(pca_result)$importance[2, ]
cumulative_variance <- cumsum(explained_variance)

# Plot cumulative variance explained
plot(cumulative_variance, xlab = "Number of Principal Components", 
     ylab = "Cumulative Variance Explained", type = "b", main = "Cumulative Variance Explained by PCs")
abline(h = 0.9, col = "red", lty = 2)  # Add a line for 90% threshold

# Number of PCs to keep (that explain 90% of cumulative variance)
n_pcs_to_keep <- which(cumulative_variance >= 0.9)[1]

# Create a new data frame with the selected PCs for training
pca_train_data <- data.frame(pca_result$x[, 1:n_pcs_to_keep])
pca_train_data$teff <- train_data$teff  # Add the target variable back

# Run linear regression on the selected PCs
pc_model <- lm(teff ~ ., data = pca_train_data)

# Prepare test data
# Scale the test predictors manually using the same means and standard deviations
test_predictors <- test_data %>% select(-teff) %>% select_if(is.numeric)

# Ensure the test predictors have the same number of columns as training predictors
print("Dimensions of test predictors:")
print(dim(test_predictors))

# Manually scale the test data using training means and standard deviations
test_predictors_scaled <- scale(test_predictors, center = training_means, scale = training_sds)

# Apply PCA transformation to the test set
pca_test_data <- data.frame(predict(pca_result, newdata = test_predictors_scaled)[, 1:n_pcs_to_keep])

# Add the response variable back for evaluation
pca_test_data$teff <- test_data$teff

# Make predictions
pca_predictions <- predict(pc_model, newdata = pca_test_data)

# Calculate MSE for the PCA model
pca_mse <- mean((pca_test_data$teff - pca_predictions)^2)
print(paste("MSE for the PCA model:", pca_mse))

# Calculate MSE for the full model for comparison
full_model <- lm(teff ~ ., data = train_data)
full_predictions <- predict(full_model, newdata = test_data)
full_mse <- mean((test_data$teff - full_predictions)^2)
print(paste("MSE for the full model:", full_mse))

# Assuming bss_mse is defined from earlier
print(paste("MSE for the best subset model:", bss_mse))
```
The MSE of the full model is still much lower than that of the PCA model and the best subset model. Removing the multicollinear variables does not help with improving predictions for this dataset. This is because the multicollinear variables have the most variance influence and greatly impacts the predictive ability of the model (as shown in the loading plot below). Using alternative modeling methods such as LASSO or Ridge regression seem to work better for this dataset. Also, transforming the data in another way other than using a log transformation might work better. Also, non-linear ML models may work better for this dataset.

### PCA Loading Plot
```{r}
# Assuming pca_result is the result from the prcomp() function
# Extract loadings
loadings <- as.data.frame(pca_result$rotation)

# Add variable names
loadings$Variable <- rownames(loadings)

# Melt the data for ggplot2
loadings_long <- loadings %>% 
  pivot_longer(-Variable, names_to = "Principal_Component", values_to = "Loading")

# Calculate the explained variance for plotting
explained_variance <- summary(pca_result)$importance[2, ]
cumulative_variance <- cumsum(explained_variance)

# Create a data frame with variable names and their explained variance
loadings_long$Variance_Explained <- rep(explained_variance, each = nrow(loadings))

# Plotting variable loadings with variance influence
p <- ggplot(loadings_long, aes(x = Variable, y = Loading, fill = Principal_Component)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  labs(title = "Variable Loadings from PCA",
       x = "Variables",
       y = "Loadings") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set3") +  # Use Set3 for more colors
  geom_text(aes(label = round(Loading, 2)), position = position_dodge(width = 0.9), vjust = -0.5)

# Print the plot
print(p)

# Save the plot
ggsave("pca_variable_loadings.png", plot = p, width = 10, height = 6)
```

